Towards practical dynamic induction control of wind farms: analysis of optimally controlled wind-farm boundary layers and sinusoidal induction control of first-row turbines

### 2018-02-06 draft.tex
Comments JM: section 2

\section{Description of optimal control simulations in MM17}\label{sec:optimal_control_sims}
The current section describes the optimal control simulations performed by MM17, the results of which are further analyzed in the current paper. First, the methodology is illustrated. Then, the numerical setup is detailed. Afterwards, the optimization results on power extraction and time-averaged flow field features are discussed.

@@

A schematic overview of the wind-farm control methodology is shown in Figure \ref{fig:optimization_meth}. Figure \ref{fig:optimization_meth}a illustrates the control block diagram: an iterative optimization loop updates the wind-farm control vector $\bs{\varphi}$ until a set of optimized controls $\bs{\varphi}^{\bullet}$ is found. This optimization is based upon an unsteady turbulence-resolving LES wind-farm flow model, and sensitivities of the cost functional $\J$ (i.e. the total wind-farm power extraction) are calculated using an adjoint formulation of this flow model. In this way, no a priori simplification on the model accuracy are made, and the control signals can be designed in a such a way that turbines can actively tap into the dynamic of the turbulent wind-farm boundary layer. The optimization is performed using a receding-horizon control framework, as illustrated in Figure \ref{fig:optimization_meth}b. In this framework, wind-farm control is optimized for a finite time horizon $T$, involving a sequential set of LES and adjoint LES simulations. Upon convergence of the optimization, optimized control signals are applied in a flow advancement simulation for a time $T_A < T$, after which a new optimization window is initiated.

@@

Within each optimization window, total wind-farm power is optimized by solving the following partial-differential-equation-constrained optimization problem: 
\begin{alignat}{4}
& \underset{\bs{\varphi}, \bs{q}}{\text{minimize}}  & \qquad  \J(\bs{\varphi}, \bs{q}) &= - \Tint \sum_{i=1}^{N_t} P_i ~\dt  & \label{eq:costfunction}\\
& \text{s.t.}                      			&         \frac{\partial \utilde}{\partial t} + \big(\utilde \cdot \nabla \big)~ \utilde &= - \nabla (\ptilde + \ptilde_\infty) / \rho - \nabla \cdot \boldsymbol{\tau}_{sgs} + \sum_{i=1}^{N_t} \bs{f}_i \qquad  & \text{in } \Omega \times (0,T], \label{eq:NSmomentum_constraint} \\
&                                                   &        \nabla \cdot \utilde&=0 									        & \text{in } \Omega \times (0,T], \label{eq:NScontinuity_constraint}\\
&                                                   &        \tau \ddt{\ctihat}&=\cti - \ctihat 								& i=1\dots N_t~\text{in } (0,T],  \label{eq:ctihat_constraint}\\
&                                                   &       0 \leq~ &\cti \leq C_{T,\text{max}}'				& i=1\dots N_t~\text{in } (0,T],  \label{eq:boxct_constraint}
\end{alignat}
The cost functional that is minimized in \eqref{eq:costfunction} is the total wind-farm energy extraction over time horizon $T$, with switched sign for convention of notation. The control variables are the time-dependent thrust coefficient setpoints $C_{T,i}'$ of every turbine $i$ ($=1 \dots N_t$), i.e. $\bs{\varphi} = [C_{T,1}'(t), \dots, C_{T,N_t}'(t) ]$, and the state variables are denoted as  $\bs{q}=[\utilde(\bs{x},t);~ \ptilde(\bs{x},t);~ \widehat{C}_{T,1}'(t), \dots, \widehat{C}_{T,N_t}'(t)]$, with $\utilde$ the filtered velocity, $\ptilde$ the filtered pressure, and $\widehat{C}_{T,i}'$ the (time-filtered) thrust coefficient for turbine $i$ (see below).

@@

The forces exerted by turbine $i$ on the boundary-layer flow are parametrized using a standard non-rotating actuator disk model as
%\begin{equation}
%$	\bs{f}_i(\bs{x},t) = - \frac{1}{2}~ \widehat{C}_{T,i}'(t)~ V_i(t)^2~ \R_i (\bs{x})~ \eperpi,$
$	\bs{f}_i(\bs{x},t) = - (1/2) \widehat{C}_{T,i}'(t) V_i(t)^2 \R_i (\bs{x}) \eperpi,$
%\end{equation}
where $\R_i$ is a smoothed representation of the geometric footprint of the turbine on the LES grid and $\eperpi$ is the rotor-perpendicular vector. Further, the disk-averaged velocity is defined as $V_i = (1/A_i) \sint \R_i(\bs{x}) \utilde \cdot \eperpi ~\dx$, with $A_i$ the rotor disk area. Mechanical power captured by the wind turbine is calculated as $P_i = (1/2) C_{P,i}'(t) V_i(t)^3 A_i$, with $C_{P,i}' = 0.9 C_{T,i}'$, resulting from a fit of LES results to momentum theory, partially eliminating the overprediction of wind-turbine power on typical wind-farm LES grid resolutions \citep{munters2017optimal}.

@@

In the remainder of this section, time-averaged wind-farm flow properties will be investigated. Here and throughout the remainder of this paper, the focus will lie on case C3t5, as it produces similar energy gains as the highest-yield case C3t0 (see Figure \ref{fig:power_MM17}a), but achieves this with smooth thrust coefficient signals. In the following discussion, the time averaging operation is denoted by an overline, and flow field variables are decomposed into mean and fluctuating components as $u = \overline{u} + u' \equiv U + u'  $. Figure \ref{fig:flowfield_statistics} illustrates time-averaged flow field quantities of the reference case (left panels, a1 -- g1) and the differences between the optimized C3t5 case and the reference case (right panels, a2 -- g2). Simulation results are averaged over the different columns and are shown as either topviews at hub height (Figures \ref{fig:flowfield_statistics}b,f,g) or sideviews through a turbine column (Figures \ref{fig:flowfield_statistics}a,c,d,e).

Figures \ref{fig:flowfield_statistics}a and \ref{fig:flowfield_statistics}b illustrate sideviews and topviews of the axial velocity throughout the wind farm. It can be seen that downstream turbines are consistently subjected to higher incoming velocities, which explains the increased energy extraction discussed above. Furthermore, a larger drop in streamwise velocity over the turbine disk can be observed, most notably in the first-row turbines, indicating deeper wakes with enhanced recovery before hitting the next row of turbines. Furthermore, it can be observed that the axial velocity in the flow above and besides the wind-turbine column is reduced, indicating that the mean-flow kinetic energy is depleted in these regions, to the benefit of the flow passing through the wind turbines.

@@

Figure~\ref{fig:flowfield_statistics}f and \ref{fig:flowfield_statistics}g show planviews at hub height of transversal turbulence and mean-flow transport of axial momemtum, i.e. $\overline{\widetilde{u}_x' \widetilde{u}_y'}$ and $\widetilde{U}_x \widetilde{U}_y$ respectively. The sign convention is such that positive values correspond to transport in the positive $y$ direction in the figure. A slight increase in turbulent transversal transport towards the wake centerline can be observed behind every row. On the other hand, mean-flow transversal momentum transport into the wake region is increased significantly behind the first two turbine rows only. Downstream of these rows, the difference between these cases and the reference case is far less coherent. The latter can be explained by the fact that it occurs in the region in which $\widetilde{U}_x$ in the intercolumn channels starts to deviate significantly from the reference case as shown in Figure~\ref{fig:flowfield_statistics}b.

The analysis of flow features given above indicate that the optimal controls in case C3t5 coherently influence the wind-farm flow field in such a way as to provide better flow conditions for downstream turbines. Increased axial velocities are observed for all downstream turbines, and enhanced momentum transport towards the turbine region is achieved. Furthermore, many of the observed flow features are most salient for the first row turbines. In the following section, the optimized thrust coefficients themselves will be investigated. It will be shown that, also from a controls perspective, first-row turbines stand out from their downstream counterparts.

>>>>

\section{Description of optimal control simulations in MM17}\label{sec:optimal_control_sims}
The current section describes the optimal control simulations performed by MM17, the results of which are further analyzed in the current paper. First, the methodology is introduced. Then, the numerical setup is detailed. Afterwards, the optimization results on power extraction and time-averaged flow field features are discussed.

@@

A schematic overview of the wind-farm control methodology is shown in Figure \ref{fig:optimization_meth}. Figure \ref{fig:optimization_meth}a illustrates the control block diagram: an iterative optimization loop updates the wind-farm control vector $\bs{\varphi}(t)$ until a set of optimized controls $\bs{\varphi}^{\bullet}(t)$ is found. This optimization is based upon an unsteady turbulence-resolving LES wind-farm flow model, and sensitivities of the cost functional $\J$ (i.e. the total wind-farm power extraction) are calculated using an adjoint formulation of this flow model. In this way, a priori simplifications to the turbulent boundary layer and wake representation are avoided as much as possible, and the control signals are designed in a such a way that turbines actively tap into the dynamic of the turbulent flow. The optimization is performed using a receding-horizon control framework, as illustrated in Figure \ref{fig:optimization_meth}b. In this framework, wind-farm controls $\bs{\varphi}(t)$ are optimized for a finite time horizon $T$, involving a sequential set of LES and adjoint LES simulations. Upon convergence of the optimization, optimized control signals are applied in a flow advancement simulation for a time $T_A < T$, after which a new optimization window is initiated.

@@

Within each optimization window, the total wind-farm power is optimized by solving the following partial-differential-equation-constrained optimization problem: 
\begin{alignat}{4}
& \underset{\bs{\varphi}, \bs{q}}{\text{minimize}}  & \qquad  \J(\bs{\varphi}, \bs{q}) &= - \Tint \sum_{i=1}^{N_t} P_i ~\dt  & \label{eq:costfunction}\\
& \text{s.t.}                      			&         \frac{\partial \utilde}{\partial t} + \big(\utilde \cdot \nabla \big)~ \utilde &= - \nabla (\ptilde + \ptilde_\infty) / \rho - \nabla \cdot \boldsymbol{\tau}_{sgs} + \sum_{i=1}^{N_t} \bs{f}_i \qquad  & \text{in } \Omega \times (0,T], \label{eq:NSmomentum_constraint} \\
&                                                   &        \nabla \cdot \utilde&=0 									        & \text{in } \Omega \times (0,T], \label{eq:NScontinuity_constraint}\\
&                                                   &        \tau \ddt{\ctihat}&=\cti - \ctihat 								& i=1\dots N_t~\text{in } (0,T],  \label{eq:ctihat_constraint}\\
&                                                   &       0 \leq~ &\cti \leq C_{T,\text{max}}'				& i=1\dots N_t~\text{in } (0,T],  \label{eq:boxct_constraint}
\end{alignat}
The cost functional that is optimized in \eqref{eq:costfunction} is the total wind-farm energy extraction over time horizon $T$. The control variables are the time-dependent thrust coefficient setpoints $C_{T,i}'$ of every turbine $i$ ($=1 \dots N_t$), i.e. $\bs{\varphi} = [C_{T,1}'(t), \dots, C_{T,N_t}'(t) ]$, and the state variables are denoted as  $\bs{q}=[\utilde(\bs{x},t);~ \ptilde(\bs{x},t);~ \widehat{C}_{T,1}'(t), \dots, \widehat{C}_{T,N_t}'(t)]$, with $\utilde$ the filtered velocity, $\ptilde$ the filtered pressure, and $\widehat{C}_{T,i}'$ the (time-filtered) thrust coefficient for turbine $i$ (see below).

@@

The forces exerted by turbine $i$ on the boundary-layer flow are parametrized using a standard non-rotating actuator disk model as
%\begin{equation}
%$	\bs{f}_i(\bs{x},t) = - \frac{1}{2}~ \widehat{C}_{T,i}'(t)~ V_i(t)^2~ \R_i (\bs{x})~ \eperpi,$
$	\bs{f}_i(\bs{x},t) = - (1/2) \widehat{C}_{T,i}'(t) V_i(t)^2 \R_i (\bs{x}) \eperpi,$
%\end{equation}
where $\R_i$ is a smoothed representation of the geometric footprint of the turbine on the LES grid and $\eperpi$ is the rotor-perpendicular vector. Further, the disk-averaged velocity is defined as $V_i = (1/A_i) \sint \R_i(\bs{x}) \utilde \cdot \eperpi ~\dx$, with $A_i$ the rotor disk area. Mechanical power captured by the wind turbine is calculated as $P_i = (1/2) C_{P,i}'(t) V_i(t)^3 A_i$, with $C_{P,i}' = 0.9 C_{T,i}'$, resulting from a fit of LES results to momentum theory, eliminating the overprediction of wind-turbine power on typical wind-farm LES grid resolutions \citep{munters2017optimal}.

@@

In the remainder of this section, time-averaged wind-farm flow properties will be investigated. Here and throughout the remainder of this paper, we focus on case C3t5, as it produces similar energy gains as the highest-yield case C3t0 (see Figure \ref{fig:power_MM17}a), but achieves this with smooth thrust coefficient signals. In the following discussion, the time averaging operation is denoted by an overline, and flow field variables are decomposed into mean and fluctuating components as $\widetilde{u} = \overline{\widetilde{u}} + \widetilde{u}' \equiv \widetilde{U} + \widetilde{u}'  $. Figure \ref{fig:flowfield_statistics} illustrates time-averaged flow field quantities of the reference case (left panels, a1 -- g1) and the differences between the optimized C3t5 case and the reference case (right panels, a2 -- g2). Simulation results are averaged over the different columns and are shown as either topviews at hub height (Figures \ref{fig:flowfield_statistics}b,f,g) or sideviews through a turbine column (Figures \ref{fig:flowfield_statistics}a,c,d,e).

Figures \ref{fig:flowfield_statistics}a and \ref{fig:flowfield_statistics}b illustrate sideviews and topviews of the axial velocity throughout the wind farm. It can be seen that downstream turbines in the controlled case experience consistently higher incoming velocities, which explains the increased energy extraction discussed above. Furthermore, a larger drop in streamwise velocity over the turbine disk can be observed, most notably in the first-row turbines, indicating deeper wakes with enhanced recovery before hitting the next row of turbines. Furthermore, it can be observed that the axial velocity in the flow above and besides the wind-turbine column is reduced, indicating that the mean-flow kinetic energy is depleted in these regions, to the benefit of the flow passing through the wind turbines.

@@

Figure~\ref{fig:flowfield_statistics}f and \ref{fig:flowfield_statistics}g show planviews at hub height of transversal turbulence and mean-flow transport of axial momemtum, i.e. $\overline{\widetilde{u}_x' \widetilde{u}_y'}$ and $\widetilde{U}_x \widetilde{U}_y$ respectively. The sign convention is such that positive values correspond to transport in the positive $y$ direction in the figure. A slight increase in turbulent transversal transport towards the wake centerline can be observed behind every row. The mean-flow transversal momentum transport into the wake region is increased significantly behind the first two turbine rows. Downstream of these rows, the difference between these cases and the reference case is far less coherent. The latter can be explained by the fact that $\widetilde{U}_x$ in the intercolumn channels starts to deviate significantly from the reference case as shown in Figure~\ref{fig:flowfield_statistics}b.

The analysis of flow features given above indicates that the optimal controls in case C3t5 influence the wind-farm flow field in such a way as to provide better flow conditions for downstream turbines. Increased axial velocities are observed for all downstream turbines, and enhanced momentum transport towards the turbine region is achieved. Furthermore, many of the observed flow features are most salient for the first row turbines. In the following section, the optimized thrust coefficients themselves will be investigated. It will be shown that, also from a controls perspective, first-row turbines stand out from their downstream counterparts.
